\LARGE \bf Learning Depth-Sensitive Conditional Random Fields for Semantic Segmentation of RGB-D Images

### 2013-09-13 paper.tex
include some remarks, add refstyle for \secref withouth paragraph symbol.
git-svn-id: https://svn.ais.uni-bonn.de/svn/staff/publications/icra_2014_mueller_nyu@4367 b2511e17-9ac0-488b-bc21-1d6d82e5baa2

Our approach starts with a random forest, providing a noisy local estimate
of semantic classes based on color and depth information. These estimates
are grouped together using a superpixel approach, for which we extend previous
superpixel algorithms from the RGB to the RGB-D domain.
We then build a geometric model of the scene, based on the neighborhood graph
of superpixels.  We use this graph not only to capture spacial relations in the
2d plane of the image, but also to model object distances and surface angles in
3d, using a point cloud generated from the RGB-D image.

We access the accuracy of our model on the challenging NYU Segmentation V2
dataset \citep{SilbermanECCV12}, where our model outperforms previous
approaches.  Our analysis shows that while our random forest model
already has competitive performance, the superpixel-based grouping and in
particular the loss-based learning are integral ingredients of the success of our method.

\section{RELATED WORK}
The task of dense semantic labeling in natural images has been widely studied
in computer vision.  Our approach of using pixel-based low-level cues is
similar to the work of \citet{shotton2006textonboost}, who introduced
TextonBoost for semantic segmentation, and used a hand-coded CRF energy to
model spacial interactions. More recent work usually employ learned conditional
random fields, learning object interactions from the
data to directly maximize a loss function~\citep{lucchi2013learning, krahenbuhl2012efficient},
which is the approach we are following.

Work using RGB-D data only recently emerged with
the availability of the NYU Depth Dataset V1~\citep{silberman2011indoor}.
\citet{silberman2011indoor} captured a wide variety of indoor scene using a structured light sensor,
and provided dense semantic object class annotation. Their work also introduced a baseline method
for semantic segmentation of RGB-D image, which used an approach based on superpixels, interest
point descriptors and a conditional random field. While pairwise potentials for the CRF were carefully
designed for the dataset, potentials were either directly set by hand or estimated using empirical
frequencies. This is in contrast to our work, which applies structured prediction techniques
to learn potentials automatically to optimize predictive performance.
\citet{ren2012rgb} focus on the design of features for semantic labeling of RGB-D data, and use
a hierarchical segmentation to provide context. While they also define a conditional random field on superpixels,
their model is again not learned, but a weighted Potts model, using only a
probability of boundary map, and not taking spacial layout into account at all.

@@

We learn the unary and pairwise energy functions $\psi_v$ and $\psi_{v, w}$ from the training data
using a structural support vector machine. 
The concept of structural support vector machines allows for a principled,
maximum-margin based, loss-sensitive training of conditional random fields, that allows much
more complex interactions than the simple Potts potentials that are often used in the literature.

@@

\subsection{Unary Image Features}\seclabel{unaries}
Our method builds on the probability output of a random forest, trained for pixel-wise
classification of the structure classes.
We use the GPU implementation provided by \citet{stueckler2013}.
The input for training are the full RGB-D images, transformed to Lab
color space. Each tree in the forest uses training pixel only from a subset of
training images. For each training image, an equal number of pixels for each
occurring class is sampled. Split features are given by difference of regions
on color or depth channels. Region size and offsets are normalized using the
depth at the target pixel.
We accumulate the probabilistic output for all pixels within a superpixel, and
use the resulting distribution as a feature for the unary node potentials in our CRF model.
We augment these prediction with another feature, based on the height of a superpixel in 3D.
This is a very informative feature, in particular to determine the floor.
To compute the height of a (super) pixel, we first find the ``up'' direction.
We use a very simple approach that we found effective: we cluster normal directions of all pixels
into 10 clusters using $k$-means, and use the one that is most parallel to the $Y$ direction, which
roughly corresponds to height in the dataset.
We then project the 3D point cloud given by the depth along this direction, and normalize the result between
$0$ and $1$.
This procedure works robustly given there is any horizontal area in the image,
such as the ground or a table. A few scenes contain only walls and furniture,
and the approach fails for these. \Figref{height} illustrates a typical case and
one of the much rarer failure cases.
While we could use a more elaborate scheme,
such as the one from \citet{SilbermanECCV12}, we suspect that the feature is of
little use in scenes without horizontal surfaces.
\begin{figure}
    \begin{center}
        \includegraphics[width=\linewidth]{images/normal_feature}
    \end{center}
    \caption{%
        Visualization of one of the pairwise features, the similarity between superpixel normals.
        The image shows the zoom-in of a bedroom scene, together with the superpixel oversegmentation.
        Lines connect adjacent superpixels, and line-strength gives the magnitude of the orientation similarity.
    \figlabel{normal_feature}}
\end{figure}

@@

    \item[Color Contrast] \hfill \\
        We employ a non-linear color contrast, as is common in the computer
        vision literature, between the superpixel mean colors $c_i$ and $c_j$:
        \[\exp\left(-\gamma \lVert c_i - c_j \rVert^2\right)\]

    \item[Vertical Alignment] \hfill \\
        We model the directed angle between superpixel centers in the 2d image
        plane.  This allows the  model to learn that ``structure'' is above
        ``floor'', but not the other way around.

    \item[Depth Difference] \hfill \\
        We include the signed depth difference between superpixels, which
        allows the model to detect depth discontinuities that are not
        represented in the 2d neighborhood graph of the superpixels.

    \item[Normal Orientations] \hfill \\
        Differences in normal vector orientation are a strong clue on
        whether two superpixels belong to the same surface, and therefore the
        same structural class.
        We compute the orientation of normals in the 3D using the method of \citet{holz_2011_robocup},
        as implemented in the point cloud library (pcl)\footnote{\url{http://pointclouds.org}}.
        All normals within a superpixel are then averaged, to get a single orientation for each superpixel.
        The feature is computed as $\frac{\pi}{4}$ minus the (undirected) angle between the normals belonging
        to two adjacent superpixels.  
        The feature is illustrated in \Figref{normal_feature}. The difference
        in normal orientation highlights that pillow and wall are distinct
        objects, even though there is no strong distinction in color or depth.
\end{description}

@@

\subsection{Results}
\Tabref{results} compares different components of our approach with the literature.
Please note that we \emph{first} designed our final model, using only the
validation data. We now report accuracies of simpler models for reference,
however these results were not used for model selection. To separate
the influence of loss-based training and the spacial reasoning of the conditional random field,
we also train a standard support vector machine on the unary potentials for comparison.

The random forest prediction, as reported in \citet{stueckler2013} is already quite competitive.
Grouping into superpixels slightly improves performance, by removing
high-frequency noise and snapping to object boundaries. Somewhat surprisingly,
using a standard unstructured SVM with rescaled loss already advances the mean
accuracy decidedly above the previous state-of-the art. We
attribute this mostly to the ability of the SVM to more directly minimize the
target loss, leading to a calibration for the actual error measure. On the
other hand, the total accuracy only decreases by $0.3\%$, while the average
accuracy increases by $4.5\%$. This means that the SVM can further exploit
uncertainty within superpixels and correlations between classes.
This performance is still improved upon, both in class average and pixel average performance
by the learned conditional random field approach, yielding the best published result
so far for both measures. The increase over the standard SVM is $1.2\%$ for class average accuracy
and $2.2\%$ for pixel average accuracy.

A visualization of the impact of each processing step can be found in
\Figref{comparison}, which shows prediction results on the test set.  The for
prediction methods correspond to the rows of \Tabref{results}.  The difference
between the SVM and CRF approaches are clearly visible, with the CRF producing
results that are very close to the ground truth in several complex scenes.

@@

\section{SUMMARY AND DISCUSSION}
We introduce a conditional random field formulation for semantic segmentation
of structure classes in RGB-D images. We base our model in the output of an efficient GPU implementation
of random forest, and model spacial neighborhood using a superpixel-based approach.
We combine color, depth and 3D orientation features into an energy function that is fully learned using 
 the structural support vector machine approach.  By explicitly modeling 3D
 relations in a fully learned framework, we improve the state-of-the art on the
 NYU V2 dataset for semantic annotation of structure classes.

While our approach allows modeling of spacial relations, these are limited to
local interactions. In future work, these interactions could be extended to
larger areas using latent variable models~\citep{dann2012pottics} or higher
order potentials~\citep{kohli2009robust}.
Another possible line of future work is to combine our approach with a more
task-specific one, directly including support plane assumptions into the model, as done by
\citet{SilbermanECCV12}.

>>>>

Our approach starts with a random forest, providing a noisy local estimate
of semantic classes based on color and depth information. These estimates
are grouped together using a superpixel approach, for which we extend previous
superpixel algorithms from the RGB to the RGB-D domain.
We then build a geometric model of the scene, based on the neighborhood graph
of superpixels.  We use this graph not only to capture spatial relations in the
2D plane of the image, but also to model object distances and surface angles in
3D, using a point cloud generated from the RGB-D image.

We assess the accuracy of our model on the challenging NYU Segmentation V2
dataset \citep{SilbermanECCV12}, where our model outperforms previous
approaches.  Our analysis shows that while our random forest model
already has competitive performance, the superpixel-based grouping and in
particular the loss-based learning are integral ingredients of the success of our method.

\section{RELATED WORK}
The task of dense semantic labeling in natural images has been widely studied
in computer vision.  Our approach of using pixel-based low-level cues is
similar to the work of \citet{shotton2006textonboost}, who introduced
TextonBoost for semantic segmentation, and used a hand-coded CRF energy to
model spatial interactions. More recent work usually employ learned conditional
random fields, learning object interactions from the
data to directly maximize a loss function~\citep{lucchi2013learning, krahenbuhl2012efficient},
which is the approach we are following.

Work using RGB-D data only recently emerged with
the availability of the NYU Depth Dataset V1~\citep{silberman2011indoor}.
\citet{silberman2011indoor} captured a wide variety of indoor scenes using a structured light sensor,
and provided dense semantic object class annotation. Their work also introduced a baseline method
for semantic segmentation of RGB-D image, which used an approach based on superpixels, interest
point descriptors and a conditional random field. While pairwise potentials for the CRF were carefully
designed for the dataset, potentials were either directly set by hand or estimated using empirical
frequencies. This is in contrast to our work, which applies structured prediction techniques
to learn potentials automatically to optimize predictive performance.
\citet{ren2012rgb} focus on the design of features for semantic labeling of RGB-D data, and use
a hierarchical segmentation to provide context. While they also define a conditional random field on superpixels,
their model is again not learned, but a weighted Potts model, using only a
probability of boundary map, and not taking spatial layout into account at all.

@@

We learn the unary and pairwise energy functions $\psi_v$ and $\psi_{v, w}$ from the training data
using a structural support vector machine~\citep{joachims2009cutting}. 
The concept of structural support vector machines allows for a principled,
maximum-margin based, loss-sensitive training of conditional random fields. Learning the potential yields much
more complex interactions than the simple Potts potentials that are often used in the literature.

@@

\subsection{Unary Image Features}\seclabel{unaries}
Our method builds on the probability output of a random forest, trained for pixel-wise
classification of the structure classes.
We use the GPU implementation provided by \citet{stueckler2013}\footnote{\url{https://github.com/deeplearningais/curfil}}.
The input for training are the full RGB-D images, transformed to Lab
color space. Each tree in the forest uses training pixels only from a subset of
training images. For each training image, an equal number of pixels for each
occurring class is sampled. Split features are given by difference of regions
on color or depth channels. Region size and offsets are normalized using the
depth at the target pixel.
We accumulate the probabilistic output for all pixels within a superpixel, and
use the resulting distribution as a feature for the unary node potentials in our CRF model.
We augment these prediction with another feature, based on the height of a superpixel in 3D.
This is a very informative feature, in particular to determine the floor.
To compute the height of a (super) pixel, we first find the ``up'' direction.
We use a very simple approach that we found effective: we cluster normal directions of all pixels
into 10 clusters using $k$-means, and use the one that is most parallel to the $Y$ direction, which
roughly corresponds to height in the dataset.
We then project the 3D point cloud given by the depth along this direction, and normalize the result between
$0$ and $1$.
This procedure works robustly given there is any horizontal area in the image,
such as the ground or a table. A few scenes contain only walls and furniture,
and the approach fails for these. \Figref{height} illustrates a typical case and
one of the much rarer failure cases.
While we could use a more elaborate scheme,
such as the one from \citet{SilbermanECCV12}, we suspect that the feature is of
little use in scenes without horizontal surfaces.
\begin{figure}
    \begin{center}
        \includegraphics[width=\linewidth]{images/normal_feature}
    \end{center}
    \caption{%
        Visualization of one of the pairwise features, the similarity between superpixel normals.
        The image shows the zoom-in of a bedroom scene, together with the superpixel over-segmentation.
        Lines connect adjacent superpixels, and line-strength gives the magnitude of the orientation similarity.
    \figlabel{normal_feature}}
\end{figure}

@@

    \item[Color Contrast] \hfill \\
        We employ a non-linear color contrast, as is common in the computer
        vision literature, between the superpixel mean colors $c_i$ and $c_j$:
        \[\exp\left(-\gamma \lVert c_i - c_j \rVert^2\right).\]

    \item[Vertical Alignment] \hfill \\
        We model the directed angle between superpixel centers in the 2D image
        plane.  This allows the  model to learn that ``structure'' is above
        ``floor'', but not the other way around.

    \item[Depth Difference] \hfill \\
        We include the signed depth difference between superpixels, which
        allows the model to detect depth discontinuities that are not
        represented in the 2D neighborhood graph of the superpixels.

    \item[Normal Orientations] \hfill \\
        Differences in normal vector orientation are a strong clue on
        whether two superpixels belong to the same surface, and therefore the
        same structural class.
        We compute the 3D orientation of normals using the method of \citet{holz_2011_robocup},
        as implemented in the point cloud library (pcl)\footnote{\url{http://pointclouds.org}}.
        All normals within a superpixel are then averaged, to get a single orientation for each superpixel.
        The feature is computed as $\frac{\pi}{4}$ minus the (undirected) angle between the normals belonging
        to two adjacent superpixels.  
        The feature is illustrated in \Figref{normal_feature}. The difference
        in normal orientation highlights that pillow and wall are distinct
        objects, even though there is no strong distinction in color or depth.
\end{description}

@@

\subsection{Results}
\Tabref{results} compares different components of our approach with the literature.
Please note that we \emph{first} designed our final model, using only the
validation data. We now report accuracies of simpler models for reference,
however these results were not used for model selection. To separate
the influence of loss-based training and the spatial reasoning of the conditional random field,
we also train a standard support vector machine on the unary potentials for comparison.

The random forest prediction, as reported in \citet{stueckler2013} is already quite competitive.
Grouping into superpixels slightly improves performance, by removing
high-frequency noise and snapping to object boundaries. Somewhat surprisingly,
using a standard unstructured SVM with rescaled loss already advances the mean
accuracy decidedly above the previous state-of-the art. We
attribute this mostly to the ability of the SVM to more directly minimize the
target loss, leading to a calibration for the actual error measure. On the
other hand, the average pixel accuracy only decreases by $0.3\%$, while the average class
accuracy increases by $4.5\%$. This means that the SVM can further exploit
uncertainty within superpixels and correlations between classes.
This performance is still improved upon, both in class average and pixel average performance
by the learned conditional random field approach, yielding the best published result
so far for both measures. The increase over the standard SVM is $1.2\%$ for class average accuracy
and $2.2\%$ for pixel average accuracy.

A visualization of the impact of each processing step can be found in
\Figref{comparison}, which shows prediction results on the test set.  The four
prediction methods correspond to the rows of \Tabref{results}.  The difference
between the SVM and CRF approaches are clearly visible, with the CRF producing
results that are very close to the ground truth in several complex scenes.

@@

\section{SUMMARY AND DISCUSSION}
We introduce a conditional random field formulation for semantic segmentation
of structure classes in RGB-D images. We base our model in the output of an efficient GPU implementation
of random forest, and model spatial neighborhood using a superpixel-based approach.
We combine color, depth and 3D orientation features into an energy function that is fully learned using 
 the structural support vector machine approach.  By explicitly modeling 3D
 relations in a fully learned framework, we improve the state-of-the art on the
 NYU V2 dataset for semantic annotation of structure classes.

While our approach allows modeling of spatial relations, these are limited to
local interactions. In future work, these interactions could be extended to
larger areas using latent variable models~\citep{dann2012pottics} or higher
order potentials~\citep{kohli2009robust}.
Another possible line of future work is to combine our approach with a more
task-specific one, directly including support plane assumptions into the model, as done by
\citet{SilbermanECCV12}.
